Legacy Data Confounds Genomics Studies

### 2019-07-13 LatexManuscript/legacy.tex
addressed some changes

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure or \sgcomment{"or" in this context is ambiguous. Do you mean "this approach is called ..." or do you mean something else?} Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

This method \sgcomment{GCAT?} identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.

@@

These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
%Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association. \sgcomment{I'm not sure what the previous sentence is saying}

@@

For the GCAT analysis, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test. 
We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
Variants flagged by the 1kGP strict mask are included in the association test and included in the FDR adjustment. 
These variants are only removed after the FDR \sgcomment{I thought we used distinct FDR For all four categories?} and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.

>>>>

To build a test statistic to represent the association across all populations simultaneously, we performed a simple logistic regression predicting genotype based on $Q$ with the logistic factor analysis (LFA) as an offset to account for population structure. This approach is called Genotype-Conditional Association Test  (GCAT) as proposed by \citep{song2015testing}. 
We also considered two alternative approaches to account for confounders, namely using the leading five  principal components, and using population membership as covariates. 
These models were broadly consistent (See Figure \ref{CompareModel}).

The GCAT method identifies a total of 24,390 variants associated to $Q$ distributed across the genome with 15,270 passing the 1kGP strict mask filter (Figures \ref{NRS_Manhattan},\ref{NRI_Manhattan}, \ref{RS_Manhattan}, and \ref{RI_Manhattan}). Most analyses below focus on the 15,270 variants satisfying the strict mask, since these variants are unlikely to be filtered by standard pipelines.
To account for the large number of tests, we used a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}. 
We tested SNPs, INDELs and repetitive regions separately as they may have different error rates (Table \ref{sigTable}).
Lists of $Q$-associated variants and individuals with low $Q$ are provided in Supplementary Data.

@@

These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
%Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Despite state-of-the-art quality controls, the variants were imputed onto genotype data, reached genome wide significance for association with biological traits, and were reported to the GWAS catalogue.
These results indicate that there is a need to remove these variants to avoid spurious GWAS associations.

@@

For the GCAT analysis, the only filtration used was the application of an minor and major allele frequency cutoff of 0.000599 (removing singletons, doubletons and tripletons) resulting in a total of S=28,516,063 variants included in the test. 
We also used the NestedRepeats mask file to flag variants inside repetitive regions as these were analyzed separately for false discovery rate estimation.
Variants flagged by the 1kGP strict mask are included in the association test and included in the four separate FDR adjustments for the four categories of variants (INDELs in Repeat and Non-Repeat regions and SNPs in Repeat and Non-Repeat regions). 
These variants are only removed after the FDR and excluded from downstream discussion of error patterns, since most population genetics analyses use the strict mask as a filter, and we expect to find problematic variants in filtered regions.
